---
title: "main_titanic"
format: html
editor: visual
---

## Required Libraries & Resources

Install the libraries and load the dataset. (Adjust path to the .csv file)

```{r}
#install.packages("skimr")
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("ggplot2")
#install.packages("tidymodels")
#install.packages("backports")
#install.packages("broom")
#install.packages("randomForest")

library(tidyverse)
library(skimr)
library(janitor)
library(ggplot2)

data_set <- read.csv("Titanic-Dataset.csv")
```

View the first 20 entries.

```{r}
data_set <- clean_names(data_set)

head(data_set)
```

Inspect the structure of the data set.

```{r}
dim(data_set)        # rows, columns
str(data_set)        # structure
```

```{r}
glimpse(data_set)    # tidy view
```

Data dictionary (additional info).

```{r}
skim(data_set)
```

The dataset contains passenger information from the RMS Titanic sinking, including variables such as survival status, passenger class, sex, age, and fare. There are several missing values in the age, cabin, and embarked columns, but no duplicate records were found.

We will use the data to analyze the factors that affected survival status among the passengers.

survived (0 = No, 1 = Yes)

Key Predictors: pclass, sex, age, sib_sp, parch, fare, embarked

------------------------------------------------------------------------

There are a lot of missing values in this dataset. No duplicates.

```{r}
cat("Missing values:\n\n")
colSums(is.na(data_set))

cat("\nNumber of duplicates:", sum(duplicated(data_set)), "\n")
```

Summary Statistics

The summary statistics show that age is spread across a wide range, indicating a diverse group of passengers. Fare is highly skewed, with most passengers paying lower amounts and a small number paying very high prices. The survival variable is imbalanced, more passengers died than survived. Because of the skewness in fare, it may need to be adjusted or capped for better analysis.

```{r}
summary(data_set)

data_set %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, median = median, sd = sd),
                   na.rm = TRUE))
```

Histograms

The age distribution appears slightly skewed. Fare is heavily right-skewed, confirming the presence of outliers. Most passengers traveled with few or no family members, as shown by the sib_sp and parch distributions. The survival count shows that more passengers died than survived. The majority of passengers were in 3rd class, and there were more males than females on board. Most passengers embarked from Southampton. These distributions suggest that class, sex, and fare may be important factors in survival.

```{r}
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 40)+ 
  labs(title = "Distribution of Age", x = "Age", y = "Count")

ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 50)+
  labs(title = "Distribution of Fare", x = "Fare", y = "Count")

ggplot(data_set, aes(x = sib_sp)) +
  geom_histogram(bins = 10, ) +
  labs(title = "Distribution of Siblings/Spouses Aboard", x = "SibSp", y = "Count")

ggplot(data_set, aes(x = parch)) +
  geom_histogram(bins = 10,) +
  labs(title = "Distribution of Parents/Children Aboard", x = "Parch", y = "Count")

ggplot(data_set, aes(x = factor(survived))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count")

ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar() +
  labs(title = "Passenger Class Distribution", x = "Pclass", y = "Count")

ggplot(data_set, aes(x = sex)) +
  geom_bar() +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")

ggplot(data_set, aes(x = embarked)) +
  geom_bar() +
  labs(title = "Port of Embarkation", x = "Embarked (C=Cherbourg, Q=Queenstown, S=Southampton)", y = "Count")
```

Boxplots

The boxplots show that passengers who survived generally paid higher fares compared to those who did not survive. The fare variable contains extreme outliers, which can distort analysis.

```{r}
ggplot(data_set, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(data_set, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

Outliers

After capping fare at the 99th percentile, the distribution becomes more stable while still keeping most of the data, making further analysis more reliable.

```{r}
#Cap fare at 99th percentile
outlier_less_ds <- data_set

upper <- quantile(outlier_less_ds$fare, 0.99, na.rm = TRUE)

outlier_less_ds$fare <- ifelse(outlier_less_ds$fare > upper, upper, outlier_less_ds$fare)


#No more outliers
ggplot(outlier_less_ds, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(outlier_less_ds, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

# WEEK 2-3: EDA, Visualization & Transformation

## Univariate Analysis — Distribution of Key Variables

The univariate analysis shows how each variable is distributed on its own. Age is somewhat normally distributed but slightly skewed, indicating a mix of younger and older passengers. Fare is heavily right-skewed, with most passengers paying low amounts and a few paying very high prices, which confirms the presence of outliers. The survival distribution shows that more passengers died than survived, so the outcome is imbalanced. Passenger class distribution shows that most people were in 3rd class, while fewer were in 1st and 2nd class. The sex distribution shows that there were more males than females on board.

```{r}
library(tidyverse)
library(ggcorrplot)

# Age distribution
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Count") +
  theme_minimal()

# Fare distribution
ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  labs(title = "Distribution of Fare", x = "Fare", y = "Count") +
  theme_minimal()

# Survival count
ggplot(data_set, aes(x = factor(survived))) +
  geom_bar(fill = "mediumseagreen", color = "white") +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count") +
  theme_minimal()

# Passenger class
ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar(fill = "mediumpurple", color = "white") +
  labs(title = "Passenger Class Distribution", x = "Class", y = "Count") +
  theme_minimal()

# Sex distribution
ggplot(data_set, aes(x = sex)) +
  geom_bar(fill = "darkorange", color = "white") +
  labs(title = "Sex Distribution", x = "Sex", y = "Count") +
  theme_minimal()
```

## Bivariate Analysis

### Correlation Analysis — Numeric Variables

The correlation analysis shows a negative relationship between passenger class and fare, meaning higher-class passengers paid more. There is also a negative relationship between passenger class and survival, indicating that higher-class passengers were more likely to survive. Most other relationships are weak, suggesting that survival is influenced by multiple variables rather than a single strong factor.

```{r}
# Select numeric columns and remove NAs
numeric_data <- data_set %>%
  select(survived, pclass, age, sib_sp, parch, fare) %>%
  drop_na()

# Correlation matrix
cor_matrix <- cor(numeric_data)

# Correlation heatmap
ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = TRUE,
           title = "Correlation Heatmap of Numeric Variables")
```

### Scatter Plot — Age vs Fare by Survival

The scatter plots show that passengers who paid higher fares were more likely to survive. Lower-class passengers had lower survival rates, reinforcing the importance of socioeconomic status. Age does not show a strong or clear relationship with survival, other than the fact that if you were in a lower class you were more likely to survive if you were younger. Overall, fare and class appear to be more important predictors than age.

```{r}
ggplot(data_set, aes(x = age, y = fare, color = factor(survived))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  labs(title = "Age vs Fare by Survival Status",
       x = "Age", y = "Fare", color = "Survival") +
  theme_minimal()

ggplot(data_set, aes(x = age, y = pclass, color = factor(survived))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  labs(title = "Age vs Class by Survival Status",
       x = "Age", y = "Class", color = "Survival") +
  theme_minimal()

ggplot(data_set, aes(x = factor(pclass), y = factor(survived), color = factor(survived))) +
  geom_jitter(alpha = 0.5, width = 0.2, height = 0.2) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  scale_y_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Passenger Class vs Survival",
       x = "Passenger Class", y = "Survival Status", color = "Survival") +
  theme_minimal()
```

### Group Comparisons — Categorical vs Numeric

The group comparisons show clear differences in survival rates. Females had a much higher survival rate than males, and passengers in higher classes were more likely to survive than those in lower classes. Fare also varies significantly by class, with higher-class passengers paying more. These patterns suggest that both gender and socioeconomic status played a major role in determining survival.

```{r}
# Survival rate by sex
ggplot(data_set, aes(x = sex, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Sex",
       x = "Sex", y = "Proportion", fill = "Survival") +
  theme_minimal()

# Survival rate by passenger class
ggplot(data_set, aes(x = factor(pclass), fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Passenger Class",
       x = "Class", y = "Proportion", fill = "Survival") +
  theme_minimal()

# Average fare by class
ggplot(data_set, aes(x = factor(pclass), y = fare, fill = factor(pclass))) +
  geom_boxplot() +
  labs(title = "Fare Distribution by Passenger Class",
       x = "Class", y = "Fare") +
  theme_minimal() +
  theme(legend.position = "none")
```

Feature Engineering

Several new variables were created to better capture passenger characteristics. FamilySize represents the total number of family members aboard, while IsAlone indicates whether a passenger was traveling alone. Deck was extracted from the cabin information, even though a lot of values were missing and labeled as unknown. GroupSize shows how many passengers shared the same ticket, and FareBin categorizes fares into levels. The cutoffs at 8, 15, and 31 roughly separate low-cost tickets from mid-range, higher, and very expensive fares based on how the data is distributed. Since fare is heavily right-skewed, most values are concentrated at the lower end, so smaller intervals are used there, while a larger range captures the high-end outliers. These new features provide more structure to the data and may improve future analysis.

```{r}
# Create FamilySize
data_set$FamilySize <- data_set$sib_sp + data_set$parch + 1

# Create IsAlone (Initialize with 0, then set 1 where FamilySize is 1)
data_set$IsAlone <- 0
data_set$IsAlone[data_set$FamilySize == 1] <- 1

data_set$Deck <- substr(data_set$cabin, 1, 1)
data_set$Deck[is.na(data_set$Deck)] <- "Unknown"

ticket_counts <- table(data_set$ticket)
data_set$GroupSize <- ticket_counts[data_set$ticket]

data_set$FareBin <- cut(data_set$fare, breaks=c(-Inf, 8, 15, 31, Inf), labels=c("Low", "Mid", "High", "Luxury"))

view(data_set)

# Bivariate Plot: Survival by FamilySize
#ggplot(data_set, aes(x = FamilySize, fill = factor(Survived))) +
#  geom_bar(position = "fill") +
#  labs(y = "Survival Rate", title = "Survival Rate by Family Size") +
#  theme_minimal()

```

New columns vs. Survival

These visualizations help confirm that the engineered features and fare bins capture meaningful patterns in the data. The fare distribution with cutoffs shows that most values are concentrated in the lower ranges, which justifies the bin choices. The survival-by-bin plot shows clear differences across fare levels, supporting the idea that fare is linked to survival. Similarly, features like FamilySize, IsAlone, Deck, and GroupSize reveal patterns that are not obvious in the raw data, making them useful for further analysis.

```{r}
ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(8, 15, 31), 
             linetype = "dashed", 
             color = "red") +
  labs(title = "Fare Distribution with Bin Cutoffs",
       x = "Fare", y = "Count") +
  theme_minimal()

ggplot(data_set, aes(x = FareBin, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Fare Bin",
       x = "Fare Bin", y = "Proportion", fill = "Survival") +
  theme_minimal()

ggplot(data_set, aes(x = FamilySize, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Family Size",
       x = "Family Size", y = "Proportion") +
  theme_minimal()

ggplot(data_set, aes(x = factor(IsAlone), fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_x_discrete(labels = c("0" = "Not Alone", "1" = "Alone")) +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate: Alone vs Not Alone",
       x = "Passenger Type", y = "Proportion") +
  theme_minimal()

ggplot(data_set, aes(x = Deck, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Deck",
       x = "Deck", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data_set, aes(x = GroupSize, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Group Size",
       x = "Group Size", y = "Proportion") +
  theme_minimal()
```

# WEEK 4: Baseline Modeling

```{r}
# WEEK 4

library(caret)

# DATA CLEANING
titanic_clean <- na.omit(data_set)

titanic_clean$survived <- as.factor(titanic_clean$survived)
titanic_clean$sex <- as.factor(titanic_clean$sex)
titanic_clean$pclass <- as.factor(titanic_clean$pclass)

# SPLIT DATASET (70/30)
set.seed(42)

index <- createDataPartition(titanic_clean$survived, p = 0.70, list = FALSE)
index <- as.vector(index) 

train_set <- titanic_clean[index, ]
test_set  <- titanic_clean[-index, ]

# BUILD MODELS
# Linear Regression (Predict Fare)
lm_model <- lm(fare ~ pclass + sex + age + sib_sp + parch, data = train_set)

# Logistic Regression (Predict Survival)
log_model <- glm(survived ~ pclass + sex + age + sib_sp + parch,
                 data = train_set, family = "binomial")

# EVALUATE LINEAR MODEL
lm_predictions <- predict(lm_model, test_set)
lm_rmse <- RMSE(lm_predictions, test_set$fare)

cat("Linear Regression RMSE:", lm_rmse, "\n")

# EVALUATE LOGISTIC MODEL
log_probs <- predict(log_model, test_set, type = "response")
log_preds <- ifelse(log_probs > 0.5, 1, 0)

conf_matrix <- confusionMatrix(
  as.factor(log_preds),
  as.factor(test_set$survived),
  positive = "1"
)

print(conf_matrix)

cat("Accuracy:", conf_matrix$overall['Accuracy'], "\n")
cat("Precision:", conf_matrix$byClass['Precision'], "\n")
cat("Recall:", conf_matrix$byClass['Recall'], "\n")

# Safe output (no crash)
cat("\nConfusion Matrix:\n")
print(conf_matrix$table)

# MODEL COEFFICIENTS
cat("\n--- Linear Regression Coefficients ---\n")
print(summary(lm_model)$coefficients)

cat("\n--- Logistic Regression Coefficients ---\n")
print(summary(log_model)$coefficients)
```

# WEEK 5: Model Evaluation and Refinement

Random Forest

```{r}
library(randomForest)
library(caret)

# ('train_set' and 'test_set' is needed from Week 4)
# Ensure NAs are gone
train_set <- na.omit(train_set)
test_set <- na.omit(test_set)

# Build the Random Forest Model
# We are predicting 'Survived' using the same features as before
set.seed(42) # For consistent results
rf_model <- randomForest(survived ~ pclass + sex + age + sib_sp + parch + fare, 
                         data = train_set, 
                         ntree = 500,        # Number of trees in the forest
                         importance = TRUE)  # To see which variables matter most

# Model Summary
print(rf_model)

# Make Predictions on the Test Set
rf_preds <- predict(rf_model, test_set)

# Evaluate Performance
# Accuracy, Precision, and Recall for Week 5
rf_conf_matrix <- confusionMatrix(rf_preds, test_set$survived, positive = "1")
print(rf_conf_matrix)

# Check Variable Importance
# This shows which features (like Sex or Age) were most important to the model
varImpPlot(rf_model, main = "Variable Importance for Titanic Survival")
```

Decision Tree

```{r}
library(caret)
library(rpart)

#remove rows with missing variables and set factors
dt_data <- data_set %>% 
  drop_na(survived, pclass, sex, age, fare, FamilySize, IsAlone)

dt_data$survived <- as.factor(dt_data$survived)
dt_data$pclass <- as.factor(dt_data$pclass)
dt_data$sex <- as.factor(dt_data$sex)

#split dataset
set.seed(42) # Ensures results are the same every time you run it
train_index <- createDataPartition(dt_data$survived, p = 0.70, list = FALSE)

#force matrix into a simple vector
train_index_vec <- as.vector(train_index)

train_set <- dt_data[train_index_vec, ]
test_set <- dt_data[-train_index_vec, ]

#decision tree
#predict if a passenger lived (1) or died (0) using cross-validation and pruning
set.seed(42)
dt_model <- train(
  survived ~ pclass + sex + age + fare + FamilySize + IsAlone, 
  data = train_set,
  method = "rpart",
  trControl = trainControl("cv", number = 10),
  tuneLength = 10 
)


#evaluate complexity parameter
plot(dt_model)
print(dt_model$bestTune)

#evaluate accuracy
dt_predictions <- predict(dt_model, newdata = test_set)
test_accuracy <- mean(dt_predictions == test_set$survived)
cat("Final Test Data Accuracy:", round(test_accuracy, 4), "\n")


#view the graph
par(xpd = NA) 

plot(dt_model$finalModel, uniform = TRUE, margin = 0.1) 
text(dt_model$finalModel, digits = 3, cex = 0.6) 

#view the text representation of the decision rules
print(dt_model$finalModel)
```

SVM

The SVM model achieved 71.83% accuracy. It is better at predicting non-survivors (Sensitivity = 0.7857) than survivors (Specificity = 0.627). The confusion matrix shows that this model is better at predicting survivors and worse at predicting non-survivors.

Sensitivity = TP / (TP + FN)

Specificity = TN / (TN + FP)

```{r}
library(caret)
library(tidyverse)

set.seed(123)

#Clean data
data_set_clean <- data_set %>%
  select(-passenger_id, -name, -ticket) %>%
  drop_na() %>%
  mutate(across(where(is.character), as.factor))

#Make the target a factor
data_set_clean$survived <- factor(data_set_clean$survived, labels = c("Yes", "No"))

#Train/Test split
train_data <- createDataPartition(data_set_clean$survived, p = 0.8, list = FALSE)
train_data <- as.vector(train_data) 

train <- data_set_clean[train_data, ]
test  <- data_set_clean[-train_data, ]

#Train Control
ctrl <- trainControl(
  method = "cv",
  number = 5
)

svm_model <- train(
  survived ~ .,
  data = train,
  method = "svmRadial",
  trControl = ctrl
)

#print(svm_model)

svm_pred <- predict(svm_model, test)

confusionMatrix(svm_pred, test$survived)
confusionMatrix(svm_pred, test$survived)$overall["Accuracy"]
```

Gradient Boosting (Doesn't work)

```{r}
library(tidyverse)
library(tidymodels)
library(xgboost)

data_clean <- data_set %>%
  select(-passenger_id, -name, -ticket) %>%
  drop_na() %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(survived = factor(survived, levels = c(0,1), labels = c("No","Yes")))

#train test split
set.seed(123)

split <- initial_split(data_clean, prop = 0.8, strata = survived)
train <- training(split)
test  <- testing(split)

#preprocessing
rec <- recipe(survived ~ ., data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

#defining model
xgb_model <- boost_tree(
  mode = "classification",
  trees = 150,
  learn_rate = 0.3,
  tree_depth = 2
) %>%
  set_engine("xgboost")

#combining into workflow
wf <- workflow() %>%
  add_model(xgb_model) %>%
  add_recipe(rec)

#fitting model
fit <- fit(wf, data = train)


#predicting and evaluating
preds <- predict(fit, test) %>%
  bind_cols(test)

conf_mat(preds, truth = survived, estimate = .pred_class)

accuracy(preds, truth = survived, estimate = .pred_class)
```

**Model Comparison:**

Performance Metrics

```{r}

library(ggplot2)
library(dplyr)

#gather test accuracies 
log_acc <- conf_matrix$overall['Accuracy']
rf_acc <- rf_conf_matrix$overall['Accuracy']
dt_acc <- test_accuracy # Created in the Decision Tree section
svm_acc <- confusionMatrix(svm_pred, test$survived)$overall["Accuracy"]

#create comparison table
model_comparison <- data.frame(
  Model = c("Logistic Regression (Baseline)", "Decision Tree", "Random Forest", "SVM"),
  Accuracy = c(log_acc, dt_acc, rf_acc, svm_acc)
)

#sort from highest to lowest
model_comparison <- model_comparison %>%
  arrange(desc(Accuracy))

cat("\n--- Final Model Comparison Table ---\n")
print(model_comparison)

#visualize
ggplot(model_comparison, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6) +
  coord_flip() + # Flips horizontally for easier reading
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  geom_text(aes(label = scales::percent(Accuracy, accuracy = 0.1)), 
            hjust = -0.2, size = 4, fontface = "bold") +
  labs(title = "Titanic Survival Prediction: Model Accuracy Comparison",
       x = "Algorithm",
       y = "Test Data Accuracy") +
  theme_minimal() +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 14))
```
