---
title: "main_titanic"
format: html
editor: visual
---

## Required Libraries & Resources

Install the libraries and load the dataset. (Adjust path to the .csv file)

```{r}
#install.packages("skimr")
#install.packages("tidyverse")
#install.packages("janitor")

library(tidyverse)
library(skimr)
library(janitor)

data_set <- read.csv("C:/Users/DZhem/Desktop/DA & PM/DA&PM Group Project/Titanic-Dataset.csv")
```

View the first 20 entries.

```{r}
data_set <- clean_names(data_set)

head(data_set)
```

Inspect the structure of the data set.

```{r}
dim(data_set)        # rows, columns
str(data_set)        # structure
```

```{r}
glimpse(data_set)    # tidy view
```

Data dictionary (additional info).

```{r}
skim(data_set)
```

We will use the data to analyze the factors that affected survival status among the passengers.

survived (0 = No, 1 = Yes)

Key Predictors: pclass, sex, age, sib_sp, parch, fare, embarked

------------------------------------------------------------------------

There are a lot of missing values in this dataset. No duplicates.

```{r}
cat("Missing values:\n\n")
colSums(is.na(data_set))

cat("\nNumber of duplicates:", sum(duplicated(data_set)), "\n")
```

Summary Statistics

```{r}
summary(data_set)

data_set %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, median = median, sd = sd),
                   na.rm = TRUE))
```

Histograms

```{r}
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 30)

ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 30)
```

Boxplots

```{r}
ggplot(data_set, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(data_set, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

Outliers

```{r}
#Cap fare at 99th percentile
outlier_less_ds <- data_set

upper <- quantile(outlier_less_ds$fare, 0.99, na.rm = TRUE)

outlier_less_ds$fare <- ifelse(outlier_less_ds$fare > upper, upper, outlier_less_ds$fare)


#No more outliers
ggplot(outlier_less_ds, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(outlier_less_ds, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```
