---
title: "main_titanic"
format: html
editor: visual
---

## Required Libraries & Resources

Install the libraries and load the dataset. (Adjust path to the .csv file)

```{r}
#install.packages("skimr")
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("ggplot2")
#install.packages("caret")

library(tidyverse)
library(skimr)
library(janitor)
library(ggplot2)

data_set <- read.csv("Titanic-Dataset.csv")
```

View the first 20 entries.

```{r}
data_set <- clean_names(data_set)

head(data_set)
```

Inspect the structure of the data set.

```{r}
dim(data_set)        # rows, columns
str(data_set)        # structure
```

```{r}
glimpse(data_set)    # tidy view
```

Data dictionary (additional info).

```{r}
skim(data_set)
```

We will use the data to analyze the factors that affected survival status among the passengers.

survived (0 = No, 1 = Yes)

Key Predictors: pclass, sex, age, sib_sp, parch, fare, embarked

------------------------------------------------------------------------

There are a lot of missing values in this dataset. No duplicates.

```{r}
cat("Missing values:\n\n")
colSums(is.na(data_set))

cat("\nNumber of duplicates:", sum(duplicated(data_set)), "\n")
```

Summary Statistics

```{r}
summary(data_set)

data_set %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, median = median, sd = sd),
                   na.rm = TRUE))
```

Histograms

```{r}
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 40)+ 
  labs(title = "Distribution of Age", x = "Age", y = "Count")

ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 50)+
  labs(title = "Distribution of Fare", x = "Fare", y = "Count")

ggplot(data_set, aes(x = sib_sp)) +
  geom_histogram(bins = 10, ) +
  labs(title = "Distribution of Siblings/Spouses Aboard", x = "SibSp", y = "Count")

ggplot(data_set, aes(x = parch)) +
  geom_histogram(bins = 10,) +
  labs(title = "Distribution of Parents/Children Aboard", x = "Parch", y = "Count")

ggplot(data_set, aes(x = factor(survived))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count")

ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar() +
  labs(title = "Passenger Class Distribution", x = "Pclass", y = "Count")

ggplot(data_set, aes(x = sex)) +
  geom_bar() +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")

ggplot(data_set, aes(x = embarked)) +
  geom_bar() +
  labs(title = "Port of Embarkation", x = "Embarked (C=Cherbourg, Q=Queenstown, S=Southampton)", y = "Count")
```

Boxplots

```{r}
ggplot(data_set, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(data_set, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

Outliers

```{r}
#Cap fare at 99th percentile
outlier_less_ds <- data_set

upper <- quantile(outlier_less_ds$fare, 0.99, na.rm = TRUE)

outlier_less_ds$fare <- ifelse(outlier_less_ds$fare > upper, upper, outlier_less_ds$fare)


#No more outliers
ggplot(outlier_less_ds, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(outlier_less_ds, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

# ── WEEK 2-3: EDA, Visualization & Transformation ──────────

## Univariate Analysis — Distribution of Key Variables

```{r}
library(tidyverse)
library(ggcorrplot)

# Age distribution
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Count") +
  theme_minimal()

# Fare distribution
ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  labs(title = "Distribution of Fare", x = "Fare", y = "Count") +
  theme_minimal()

# Survival count
ggplot(data_set, aes(x = factor(survived))) +
  geom_bar(fill = "mediumseagreen", color = "white") +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count") +
  theme_minimal()

# Passenger class
ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar(fill = "mediumpurple", color = "white") +
  labs(title = "Passenger Class Distribution", x = "Class", y = "Count") +
  theme_minimal()

# Sex distribution
ggplot(data_set, aes(x = sex)) +
  geom_bar(fill = "darkorange", color = "white") +
  labs(title = "Sex Distribution", x = "Sex", y = "Count") +
  theme_minimal()
```

## Bivariate Analysis

### Correlation Analysis — Numeric Variables

```{r}
# Select numeric columns and remove NAs
numeric_data <- data_set %>%
  select(survived, pclass, age, sib_sp, parch, fare) %>%
  drop_na()

# Correlation matrix
cor_matrix <- cor(numeric_data)

# Correlation heatmap
ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = TRUE,
           title = "Correlation Heatmap of Numeric Variables")
```

### Scatter Plot — Age vs Fare by Survival

```{r}
ggplot(data_set, aes(x = age, y = fare, color = factor(survived))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  labs(title = "Age vs Fare by Survival Status",
       x = "Age", y = "Fare", color = "Survival") +
  theme_minimal()

ggplot(data_set, aes(x = age, y = pclass, color = factor(survived))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  labs(title = "Age vs Class by Survival Status",
       x = "Age", y = "Class", color = "Survival") +
  theme_minimal()

ggplot(data_set, aes(x = factor(pclass), y = factor(survived), color = factor(survived))) +
  geom_jitter(alpha = 0.5, width = 0.2, height = 0.2) +
  scale_color_manual(values = c("0" = "tomato", "1" = "steelblue"),
                     labels = c("Died", "Survived")) +
  scale_y_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Passenger Class vs Survival",
       x = "Passenger Class", y = "Survival Status", color = "Survival") +
  theme_minimal()
```

### Group Comparisons — Categorical vs Numeric

```{r}
# Survival rate by sex
ggplot(data_set, aes(x = sex, fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Sex",
       x = "Sex", y = "Proportion", fill = "Survival") +
  theme_minimal()

# Survival rate by passenger class
ggplot(data_set, aes(x = factor(pclass), fill = factor(survived))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "tomato", "1" = "steelblue"),
                    labels = c("Died", "Survived")) +
  labs(title = "Survival Rate by Passenger Class",
       x = "Class", y = "Proportion", fill = "Survival") +
  theme_minimal()

# Average fare by class
ggplot(data_set, aes(x = factor(pclass), y = fare, fill = factor(pclass))) +
  geom_boxplot() +
  labs(title = "Fare Distribution by Passenger Class",
       x = "Class", y = "Fare") +
  theme_minimal() +
  theme(legend.position = "none")
```

Feature Engineering

```{r}
# Create FamilySize
data_set$FamilySize <- data_set$sib_sp + data_set$parch + 1

# Create IsAlone (Initialize with 0, then set 1 where FamilySize is 1)
data_set$IsAlone <- 0
data_set$IsAlone[data_set$FamilySize == 1] <- 1

data_set$Deck <- substr(data_set$cabin, 1, 1)
data_set$Deck[is.na(data_set$Deck)] <- "Unknown"

ticket_counts <- table(data_set$ticket)
data_set$GroupSize <- ticket_counts[data_set$ticket]

data_set$FareBin <- cut(data_set$fare, breaks=c(-Inf, 8, 15, 31, Inf), labels=c("Low", "Mid", "High", "Luxury"))

view(data_set)

# Bivariate Plot: Survival by FamilySize
#ggplot(data_set, aes(x = FamilySize, fill = factor(Survived))) +
#  geom_bar(position = "fill") +
#  labs(y = "Survival Rate", title = "Survival Rate by Family Size") +
#  theme_minimal()

```

# WEEK 5

```{r}
library(caret)
library(rpart)

#DATA PREPARATION (Required for Advanced Modeling)
#Remove rows with missing values (NA) and set factors
dt_data <- data_set %>% 
  drop_na(survived, pclass, sex, age, fare, FamilySize, IsAlone)

dt_data$survived <- as.factor(dt_data$survived)
dt_data$pclass <- as.factor(dt_data$pclass)
dt_data$sex <- as.factor(dt_data$sex)

#SPLIT DATASET (70% Train/30% Test)
set.seed(42) # Ensures results are the same every time you run it
train_index <- createDataPartition(dt_data$survived, p = 0.70, list = FALSE)

# Force the matrix into a simple vector 
train_index_vec <- as.vector(train_index)

train_set <- dt_data[train_index_vec, ]
test_set <- dt_data[-train_index_vec, ]


#BUILD ADVANCED MODELS
#Decision Tree (Classification: Predicting 'survived')
#Predict if a passenger lived (1) or died (0) using cross-validation and pruning
set.seed(42)
dt_model <- train(
  survived ~ pclass + sex + age + fare + FamilySize + IsAlone, 
  data = train_set,
  method = "rpart",
  trControl = trainControl("cv", number = 10),
  tuneLength = 10 
)


#EVALUATE MODELS
#Evaluate Complexity Parameter (cp)
plot(dt_model)
print(dt_model$bestTune)

#Evaluate Decision Tree Accuracy
dt_predictions <- predict(dt_model, newdata = test_set)
test_accuracy <- mean(dt_predictions == test_set$survived)
cat("Final Test Data Accuracy:", round(test_accuracy, 4), "\n")


#INTERPRET MODEL RULES
#View the graphical representation of the decision rules
par(xpd = NA) 
# Added margin = 0.1 to give the tree breathing room from the edges
plot(dt_model$finalModel, margin = 0.1) 
# Added cex = 0.6 to shrink the text to 60% of its normal size so it stops overlapping
text(dt_model$finalModel, digits = 3, cex = 0.6) 

#View the text representation of the decision rules
print(dt_model$finalModel)
print(dt_model$finalModel)
```
