---
title: "Project"
format: html
editor: visual
---

## Required Libraries & Resources

Install the libraries and load the dataset. (Adjust path to the .csv file)

```{r}
#install.packages("skimr")
#install.packages("tidyverse")
#install.packages("janitor")

library(tidyverse)
library(skimr)
library(janitor)

data_set <- read.csv("Titanic-Dataset.csv")
```

View the first 20 entries.

```{r}
data_set <- clean_names(data_set)

head(data_set)
```

Inspect the structure of the data set.

```{r}
dim(data_set)        # rows, columns
str(data_set)        # structure
```

```{r}
glimpse(data_set)    # tidy view
```

Data dictionary (additional info).

```{r}
skim(data_set)
```
| Variable | Type | Description | Notes |
|---|---|---|---|
| passenger_id | integer | Unique ID for each passenger | Not a predictor |
| survived | integer | Survival status | 0 = Died, 1 = Survived — **target variable** |
| pclass | integer | Ticket class | 1 = 1st, 2 = 2nd, 3 = 3rd — proxy for wealth |
| name | character | Full passenger name | Not used in analysis |
| sex | character | Passenger sex | male / female |
| age | numeric | Age in years | 177 missing values |
| sib_sp | integer | Siblings/spouses aboard | 0 = traveled alone |
| parch | integer | Parents/children aboard | 0 = traveled alone |
| ticket | character | Ticket number | Not used in analysis |
| fare | numeric | Ticket price paid | Right-skewed, outliers capped |
| cabin | character | Cabin number | Mostly missing, not used |
| embarked | character | Port of embarkation | C = Cherbourg, Q = Queenstown, S = Southampton |

We will use the data to analyze the factors that affected survival status among the passengers.

survived (0 = No, 1 = Yes)

Key Predictors: pclass, sex, age, sib_sp, parch, fare, embarked

------------------------------------------------------------------------

There are a lot of missing values in this dataset. No duplicates.

```{r}
cat("Missing values:\n\n")
colSums(is.na(data_set))

cat("\nNumber of duplicates:", sum(duplicated(data_set)), "\n")
```

Summary Statistics

```{r}
summary(data_set)

data_set %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, median = median, sd = sd),
                   na.rm = TRUE))
```

Histograms

```{r}
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 40)+ 
  labs(title = "Distribution of Age", x = "Age", y = "Count")

ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 50)+
  labs(title = "Distribution of Fare", x = "Fare", y = "Count")

ggplot(data_set, aes(x = sib_sp)) +
  geom_histogram(bins = 10, ) +
  labs(title = "Distribution of Siblings/Spouses Aboard", x = "SibSp", y = "Count")

ggplot(data_set, aes(x = parch)) +
  geom_histogram(bins = 10,) +
  labs(title = "Distribution of Parents/Children Aboard", x = "Parch", y = "Count")

ggplot(data_set, aes(x = factor(survived))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count")

ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar() +
  labs(title = "Passenger Class Distribution", x = "Pclass", y = "Count")

ggplot(data_set, aes(x = sex)) +
  geom_bar() +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")

ggplot(data_set, aes(x = embarked)) +
  geom_bar() +
  labs(title = "Port of Embarkation", x = "Embarked (C=Cherbourg, Q=Queenstown, S=Southampton)", y = "Count")
```

Boxplots

```{r}
ggplot(data_set, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(data_set, aes(x = factor(survived), y = fare)) +
  geom_boxplot()


```

Outliers

```{r}
#Cap fare at 99th percentile
outlier_less_ds <- data_set

upper <- quantile(outlier_less_ds$fare, 0.99, na.rm = TRUE)

outlier_less_ds$fare <- ifelse(outlier_less_ds$fare > upper, upper, outlier_less_ds$fare)


#No more outliers
ggplot(outlier_less_ds, aes(x = factor(survived), y = age)) +
  geom_boxplot()

ggplot(outlier_less_ds, aes(x = factor(survived), y = fare)) +
  geom_boxplot()
```

# ── WEEK 2-3: EDA, Visualization & Transformation ──────────

## Univariate Analysis — Distribution of Key Variables

```{r}
library(tidyverse)
library(ggcorrplot)

# Age distribution
ggplot(data_set, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Count") +
  theme_minimal()

# Fare distribution
ggplot(data_set, aes(x = fare)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  labs(title = "Distribution of Fare", x = "Fare", y = "Count") +
  theme_minimal()

# Survival count
ggplot(data_set, aes(x = factor(survived))) +
  geom_bar(fill = "mediumseagreen", color = "white") +
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) +
  labs(title = "Survival Count", x = "Survived", y = "Count") +
  theme_minimal()

# Passenger class
ggplot(data_set, aes(x = factor(pclass))) +
  geom_bar(fill = "mediumpurple", color = "white") +
  labs(title = "Passenger Class Distribution", x = "Class", y = "Count") +
  theme_minimal()

# Sex distribution
ggplot(data_set, aes(x = sex)) +
  geom_bar(fill = "darkorange", color = "white") +
  labs(title = "Sex Distribution", x = "Sex", y = "Count") +
  theme_minimal()
```